---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# geozarr

<!-- badges: start -->
<!-- badges: end -->

The `geozarr` package implements a set of conventions for geospatial data on top of the Zarr specification. It builds on the `zarr` package, a native R implementation of the Zarr specification that can read and write Zarr v.3 stores in memory, on the local file system and over HTTP.

The following conventions are supported by `geozarr`:

- [cs](https://github.com/R-CF/zarr_conventions_cs): Comprehensive support for any kind of axis, with CF-compatible constructs.
- [spatial](https://github.com/zarr-conventions/spatial): Compact coordinate system for X-Y (image, GIS) data.
- [proj](https://github.com/zarr-conventions/proj): Reference frame to register a coordinate system to Earth.
- [uom](https://github.com/clbarnes/zarr-convention-uom): Unit-of-measure information for data in a Zarr array.
- [ref](https://github.com/R-CF/zarr_convention_ref): A standard way to refer to Zarr objects or attributes elsewhere in the store or in other stores.

# Working with `geozarr`

The `geozarr` package is closely integrated with the `zarr` package, to the extent that the only user-facing function in this package is `as_geozarr()`, to convert an R object (vector, matrix, array) into a Zarr array or store with GeoZarr metadata. Manipulating the Zarr object is done with the same tools as a regular Zarr object.

The `as_geozarr() function creates a GeoZarr object from an R matrix or array. A GeoZarr object is like a Zarr object but with special attributes to establish a coordinate system. Default settings will be taken from the R object (data type, shape). Data is chunked into chunks of length 100 (or less if the array is smaller) and compressed. The object may be a stand-alone Zarr store (single Zarr array only), or a Zarr store to which additional Zarr groups and arrays may be added. The Zarr store may be in memory or persisted to a local file system.

Depending on the properties of the R object, the GeoZarr object may use the "spatial" or "cs" convention for encoding. The "spatial" encoding is the most compact and it will be used for R objects that have at least X and Y dimensions, identified by the names set on the dimensions, and an optional third axis which is typically an image band or a discrete (class) axis -- the third axis may not represent height/depth (Z) or time (T). The coordinates must be numeric and regularly spaced and the Y coordinates must be decreasing. In other words, the "spatial" convention will be used for imagery style, north-up arrays with a coordinate system tied to the top-left corner of the array space. For all other cases the "cs" convention will be used which can use any type and number of axes, including Z and T.

If the coordinates along the axes (the `dimnames` of the R object) are not regularly spaced, secondary Zarr arrays will be created with the axis coordinates, if the length of the axis is longer than the option `GeoZarr.options$max_explicit` -- shorter sets of coordinates are stored in the Zarr array `cs` attributes.

Any time coordinates will be converted to a `CFtime` format with a reference of "days since 1970-01-01", compatible with the standard system clock.

``` {r as_geozarr}
library(geozarr)

# Create an R array
x <- array(1:400, c(5, 20, 4))

# `spatial` convention
# Set named dim_names with `x` and `y` (decreasing values), third dimension is class-based
dimnames(x) <- list(x = 100000 + 0:4 * 10000, y = 19:0 * 5000, cls = letters[1:4])
z <- as_geozarr(x, "spatial_data")
z[["/spatial_data"]]
z$hierarchy()

# `cs` convention
# Named dimensions with `y` coordinates in natural order, third dimension is time (regular)
dimnames(x) <- list(x = 100000 + 0:4 * 10000, y = 0:19 * 5000, time = sprintf("2026-06-%02d", 1:4))
z <- as_geozarr(x, "cs_data")
z[["/cs_data"]]
z$hierarchy()

# Irregular time dimension in months: 31, 28, 31 and 30 days
old_explicit <- geozarr_options()$max_explicit
geozarr_options("max_explicit", 3L) # Force writing of external Zarr array
dimnames(x) <- list(x = 100000 + 0:4 * 10000, y = 0:19 * 5000, time = sprintf("2026-%02d-01", 1:4))
z <- as_geozarr(x, "cs_data_irregular_time")
z[["/cs_data_irregular_time"]]
z$hierarchy()
geozarr_options("max_explicit", old_explicit)
```

# Development

GeoZarr is currently under active development and this package is similarly in flux. The conventions implemented in this package will remain available unless the convention is deprecated due to any reason that would recommend against continuing to use the convention.

This package should currently not be used for production environments. Things may fail and you are advised to ensure that you have backups of all data that you put in a Zarr store with this package.

Like GeoZarr itself, this package is modular and allows for additional conventions to be added to this basic implementation. If you have specific needs, open an [issue on Github](https://github.com/R-CF/geozarr/issues) or, better yet, fork the code and submit code suggestions via a pull request. Specific guidance for developers is being drafted.

Installation from CRAN of the latest release:  
```
install.packages("geozarr")
```
You can install the development version of `geozarr` from [GitHub](https://github.com/R-CF/geozarr) with:
```
# install.packages("devtools")
devtools::install_github("R-CF/geozarr")
```
